Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-24 optim.qmd
edit introduction

-   British English ("American or British usage is accepted, but not a
    mixture of these")\*

@@

    -   Use jellyfish search optimiser, or JSO
    -   "hyper-parameter" rather than "hyperparameter"
    -   PP index rather than PPI

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *PP index*
(@hall1989polynomial, @cook1993projection, @lee2010projection,
@Loperfido2018, @Loperfido2020). Given high-dimensional data
$X \in \mathbf{R}^{n\times p}$ and the index function $f(\cdot)$, PP
finds the orthonormal projection matrix $A \in \mathbf{R}^{p \times d}$
by solving the following optimisation problem:

@@

These indices $f(\cdot)$ are characterised by the informativeness or
"interestingness" of a projection and are often non-linear and
non-convex. As such, an effective and efficient optimisation procedure
is essential to enable sufficient exploration of the data landscape
during the visualisation process and to arrive at a global optimal
viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP and facilitating a better experience in the visualisation process.
@cook1995grand introduced the PP guided tour, which enabled interactive
visualisation of the optimisation to visually explore high-dimensional
data. It is implemented in the R [@R] package `tourr`[@tourr].
@RJ-2021-105 highlighted potential problems of the optimisation routine
implemented. While improving the quality of the optimisation solutions
in the tour is essential, it is also important to be able to watch the
projected data as the optimisation progresses. As such, it is pertinent
to integrate the guided tour with a global optimisation algorithm that
is efficient in finding the global optimal and enables viewing of the
projected data during the exploration process.

The artificial Jellyfish Search Optimiser (JSO) [@chou_novel_2021] is a
swarm-based metaheuristic designed to solve global optimisation
problems. Inspired by the search behaviour of jellyfish in the ocean,
JSO is one of the latest swarm intelligence algorithms
[@rajwar_exhaustive_2023], which was shown to have stronger search
ability and faster convergence with fewer parameters compared to classic
optimisation methods [@chou_novel_2021]-[@chou_recent_2022]. These
practical properties makes JSO a strong candidate in integration with
the PP guided tour. As such, it is of interest to explore the potential
of JSO in enhancing the PP guided tour and examine the characteristics
and advantages of such implementation.

The primary goal of this study is to investigate the performance of JSO
in the context of the Projection pursuit. A series of simulation
experiments using well-recognised data sets and PP indexes are conducted
to yield insights regarding the behaviour of JSO under different
settings and the sensitivity of hyper-parameter in the optimisation.
Second, to observe the performance of JSO with different types of PP
indexes, a collection of metrics is introduced to capture specific
properties of the index including squintability and smoothness as
introduced in @laa_using_2020. To the best of the authors' knowledge,
this is the first attempt to quantitatively measure squintability and
smoothness in this context. Finally, the relationship between the JSO
performance, hyper-parameters tuning and various properties of PP
indexes is analysed to provide helpful guidance for practitioners that
are using the guided tour for high dimensional visualisation.

The rest of this paper is structured as follows. @sec-background
introduces the background of the PP guided tour and reviews existing
methods in the literature. @sec-theory describes the details of JSO and
introduces metrics that measure different properties of PP indexes.
@sec-vis visualises the behaviour of the JSO in terms of describes a
simulation study on the performance of the JSO using well-known data
sets and index functions. @sec-app displays two sets of simulation
experiments comparing the JSO with the search-better optimiser and
studies the impact of different PP index properties on the optimisation
performance. @sec-conclusion summarises the work and provides
suggestions for future directions.

>>>>

-   British English ("American or British usage is accepted, but not a
    mixture of these")

@@

    -   Use jellyfish search optimiser, or JSO
    -   "hyper-parameter" rather than "hyperparameter"
    -   PP index rather than PPI, plural PP indexes
    -   datasets rather than data sets

@@

Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction technique 
aimed at identifying informative linear projections of data. The method involves 
optimising a objective function known as PP index
(@hall1989polynomial, @cook1993projection, @lee2010projection,
@Loperfido2018, @Loperfido2020), which defines the criteria for what constitutes  interesting or informative projections. Given high-dimensional data
$X \in \mathbf{R}^{n\times p}$ and an index function $f(\cdot)$, PP
finds the orthonormal projection matrix $A \in \mathbf{R}^{p \times d}$ that 
maximises the index value of the projection, $Y = XA$:

@@

The index functions can be non-linear and non-convex, hence an 
effective and efficient optimisation procedure is essential to 
explore the data landscape and achieving a globally optimal viewpoint 
of the data.

Significant efforts have been devoted to enhancing the functionality of
PP and facilitating a better experience in the visualisation process.
@cook1995grand introduces the PP guided tour, which enabled interactive 
visualisation of the optimisation in order to visually explore the high-dimensional data. It is implemented in the R [@R] package `tourr` [@tourr].
@RJ-2021-105 highlighted potential problems of the optimisation routine
implemented. While improving the quality of the optimisation solutions
in the tour is essential, it is also important to be able to watch the
projected data as the optimisation progresses. As such, it is pertinent
to integrate the guided tour with a global optimisation algorithm that
is efficient in finding the global optimal and enables viewing of the
projected data during the exploration process.

This study explores the potential for a Jellyfish Search Optimiser (JSO) to be integrated with projection 
pursuit guided tour. JSO [@chou_novel_2021] -[@rajwar_exhaustive_2023], inspired by the search behaviour of jellyfish 
in the ocean, is a swarm-based metaheuristic designed to solve global
optimisation problems. Compared to traditional methods, JSO 
demonstrated stronger search ability and faster convergence, and 
requires fewer tuning parameters. These
practical advantages make JSO as a promising candidate for enhancing
PP optimisation. As such, the study aims to explore JSO's potential
in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.

The primary goal of this study is to investigate the performance of JSO
in PP optimisation. A series of simulation
experiments using well-recognised datasets and various PP indexes are conducted
to yield insights regarding the behaviour of JSO under different
settings and its sensitivity to hyper-parameter.
Second, to observe the performance of JSO with different types of PP
indexes, metrics are introduced to capture specific
properties of the index including squintability and smoothness as
introduced in @laa_using_2020. To the best of the authors' knowledge,
this is the first attempt to quantitatively measure squintability and
smoothness within this context. Finally, the relationship between the JSO
performance, hyper-parameters tuning and various properties of PP
indexes is analysed to provide helpful guidance for practitioners using projection pursuit.

The rest of this paper is structured as follows. @sec-background
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature. @sec-theory details JSO and
introduces metrics that measure different properties of PP indexes.
@sec-vis outlines the data structures, numerical summaries, and visualisations used in the simulation studies conducted in @sec-app to assess JSO's perforamnce.
@sec-app presents two sets of simulation
experiments, comparing JSO with another optimiser, search better, and
studies the impact of different PP index properties on the optimisation
performance. @sec-conclusion summarises the work and provides
suggestions for future directions.

### 2024-06-24 optim.tex
edit introduction

\begin{itemize}
\item
  British English (``American or British usage is accepted, but not a
  mixture of these'')*
\item
  No ``we'' - always third person
\item
  Check if the affiliation information is correct
\item
  word use:

  \begin{itemize}
  \tightlist
  \item
    Use jellyfish search optimiser, or JSO
  \item
    ``hyper-parameter'' rather than ``hyperparameter''
  \item
    PP index rather than PPI
  \end{itemize}
\end{itemize}

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely \emph{PP index}
(\citet{hall1989polynomial}, \citet{cook1993projection},
\citet{lee2010projection}, \citet{Loperfido2018},
\citet{Loperfido2020}). Given high-dimensional data
\(X \in \mathbf{R}^{n\times p}\) and the index function \(f(\cdot)\), PP
finds the orthonormal projection matrix
\(A \in \mathbf{R}^{p \times d}\) by solving the following optimisation
problem:

@@

These indices \(f(\cdot)\) are characterised by the informativeness or
``interestingness'' of a projection and are often non-linear and
non-convex. As such, an effective and efficient optimisation procedure
is essential to enable sufficient exploration of the data landscape
during the visualisation process and to arrive at a global optimal
viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP and facilitating a better experience in the visualisation process.
\citet{cook1995grand} introduced the PP guided tour, which enabled
interactive visualisation of the optimisation to visually explore
high-dimensional data. It is implemented in the R \citep{R} package
\texttt{tourr}\citep{tourr}. \citet{RJ-2021-105} highlighted potential
problems of the optimisation routine implemented. While improving the
quality of the optimisation solutions in the tour is essential, it is
also important to be able to watch the projected data as the
optimisation progresses. As such, it is pertinent to integrate the
guided tour with a global optimisation algorithm that is efficient in
finding the global optimal and enables viewing of the projected data
during the exploration process.

The artificial Jellyfish Search Optimiser (JSO) \citep{chou_novel_2021}
is a swarm-based metaheuristic designed to solve global optimisation
problems. Inspired by the search behaviour of jellyfish in the ocean,
JSO is one of the latest swarm intelligence algorithms
\citep{rajwar_exhaustive_2023}, which was shown to have stronger search
ability and faster convergence with fewer parameters compared to classic
optimisation methods \citep{chou_novel_2021}-\citep{chou_recent_2022}.
These practical properties makes JSO a strong candidate in integration
with the PP guided tour. As such, it is of interest to explore the
potential of JSO in enhancing the PP guided tour and examine the
characteristics and advantages of such implementation.

The primary goal of this study is to investigate the performance of JSO
in the context of the Projection pursuit. A series of simulation
experiments using well-recognised data sets and PP indexes are conducted
to yield insights regarding the behaviour of JSO under different
settings and the sensitivity of hyper-parameter in the optimisation.
Second, to observe the performance of JSO with different types of PP
indexes, a collection of metrics is introduced to capture specific
properties of the index including squintability and smoothness as
introduced in \citet{laa_using_2020}. To the best of the authors'
knowledge, this is the first attempt to quantitatively measure
squintability and smoothness in this context. Finally, the relationship
between the JSO performance, hyper-parameters tuning and various
properties of PP indexes is analysed to provide helpful guidance for
practitioners that are using the guided tour for high dimensional
visualisation.

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the background of the PP guided
tour and reviews existing methods in the literature.
Section~\ref{sec-theory} describes the details of JSO and introduces
metrics that measure different properties of PP indexes.
Section~\ref{sec-vis} visualises the behaviour of the JSO in terms of
describes a simulation study on the performance of the JSO using
well-known data sets and index functions. Section~\ref{sec-app} displays
two sets of simulation experiments comparing the JSO with the
search-better optimiser and studies the impact of different PP index
properties on the optimisation performance. Section~\ref{sec-conclusion}
summarises the work and provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, rightrule=.15mm, bottomrule=.15mm, colback=white, bottomtitle=1mm, arc=.35mm, opacityback=0, opacitybacktitle=0.6, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, breakable, coltitle=black, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}]

>>>>

\begin{itemize}
\item
  British English (``American or British usage is accepted, but not a
  mixture of these'')
\item
  No ``we'' - always third person
\item
  Check if the affiliation information is correct
\item
  word use:

  \begin{itemize}
  \tightlist
  \item
    Use jellyfish search optimiser, or JSO
  \item
    ``hyper-parameter'' rather than ``hyperparameter''
  \item
    PP index rather than PPI, plural PP indexes
  \item
    datasets rather than data sets
  \end{itemize}
\end{itemize}

@@

Projection Pursuit (PP) (\citet{kr69}, \citet{FT74}) is a dimension
reduction technique aimed at identifying informative linear projections
of data. The method involves optimising a objective function known as PP
index (\citet{hall1989polynomial}, \citet{cook1993projection},
\citet{lee2010projection}, \citet{Loperfido2018},
\citet{Loperfido2020}), which defines the criteria for what constitutes
interesting or informative projections. Given high-dimensional data
\(X \in \mathbf{R}^{n\times p}\) and an index function \(f(\cdot)\), PP
finds the orthonormal projection matrix
\(A \in \mathbf{R}^{p \times d}\) that maximises the index value of the
projection, \(Y = XA\):

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieving a globally optimal viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP and facilitating a better experience in the visualisation process.
\citet{cook1995grand} introduces the PP guided tour, which enabled
interactive visualisation of the optimisation in order to visually
explore the high-dimensional data. It is implemented in the R \citep{R}
package \texttt{tourr} \citep{tourr}. \citet{RJ-2021-105} highlighted
potential problems of the optimisation routine implemented. While
improving the quality of the optimisation solutions in the tour is
essential, it is also important to be able to watch the projected data
as the optimisation progresses. As such, it is pertinent to integrate
the guided tour with a global optimisation algorithm that is efficient
in finding the global optimal and enables viewing of the projected data
during the exploration process.

This study explores the potential for a Jellyfish Search Optimiser (JSO)
\citep{chou_novel_2021} -\citep{chou_recent_2022} to be integrated with
projection pursuit guided tour. JSO, inspired by the search behaviour of
jellyfish in the ocean, is a swarm-based metaheuristic designed to solve
global optimisation problems. Compared to traditional methods, JSO
demonstrated stronger search ability and faster convergence, and
requires fewer tuning parameters. These practical advantages make JSO as
a promising candidate for enhancing PP optimisation. As such, the study
aims to explore JSO's potential in enhancing projection pursuit and
examines the characteristics and advantages of such implementation.

The primary goal of this study is to investigate the performance of JSO
in PP optimisation. A series of simulation experiments using
well-recognised datasets and various PP indexes are conducted to yield
insights regarding the behaviour of JSO under different settings and its
sensitivity to hyper-parameter. Second, to observe the performance of
JSO with different types of PP indexes, metrics are introduced to
capture specific properties of the index including squintability and
smoothness as introduced in \citet{laa_using_2020}. To the best of the
authors' knowledge, this is the first attempt to quantitatively measure
squintability and smoothness within this context. Finally, the
relationship between the JSO performance, hyper-parameters tuning and
various properties of PP indexes is analysed to provide helpful guidance
for practitioners using projection pursuit.

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the background of PP guided tour
and reviews existing optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-vis} outlines the
data structures, numerical summaries, and visualisations used in the
simulation studies conducted in Section~\ref{sec-app} to assess JSO's
perforamnce. Section~\ref{sec-app} presents two sets of simulation
experiments, comparing JSO with another optimiser, search better, and
studies the impact of different PP index properties on the optimisation
performance. Section~\ref{sec-conclusion} summarises the work and
provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, breakable, bottomrule=.15mm, toprule=.15mm, opacityback=0, titlerule=0mm, coltitle=black, bottomtitle=1mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, arc=.35mm, colback=white, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white]
